They're on Fire: Team Momentum in Round-Based First Person Shooter Esports

### 2022-07-30 presentation.qmd
update presentation with nhl, nba, mlb table and analysis

```{r}
#| label: setup-eval
library(tidyverse)
library(extrafont)
library(ggtext)
library(ggsci)
library(ggnewscale)
library(magick)

@@

-   2020-2022 Call of Duty League (CDL) matches (qualifiers and majors)
-   4v4.
-   First to 6 round wins
-   Teams swap sides (offense and defense) after every round
:::

@@

cod_max_diff_win_series_prop <- max(abs(cod_o_win_prop$diff_win_series_prop))
p_cod_o_series_win_prop_over_expected <- cod_o_win_prop |> 
  ggplot() +
  common_cod_heatmap_layers() +
  heatmap_tile(
    .fill = 'diff_win_series_prop'
  ) +
  heatmap_text_layers(
    .df = cod_o_win_prop, 
    .f = o_series_win_over_expected_text_layer,
    .threshold = 0.08,
    .labeller = cod_heatmap_div_labeller
  ) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('brown'),
    high = extract_ggsci_mid_color('cyan'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -cod_max_diff_win_series_prop, 
      +cod_max_diff_win_series_prop
    )
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_cod_o_series_win_prop_over_expected
save_plot_for_slide(p_cod_o_series_win_prop_over_expected)
```

@@

val_max_diff_win_series_prop <- max(abs(val_o_win_prop$diff_win_series_prop))
p_val_o_series_win_prop_over_expected <- val_o_win_prop |> 
  ggplot() +
  common_val_heatmap_layers() +
  heatmap_tile(
    .fill = 'diff_win_series_prop'
  ) +
  heatmap_text_layers(
    .df = val_o_win_prop, 
    .f = o_series_win_over_expected_text_layer,
    .threshold = 0.2,
    .labeller = val_heatmap_div_labeller
  ) +
  val_swap_path() +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('brown'),
    high = extract_ggsci_mid_color('cyan'),
    mid = 'white',
    limits = c(
      -val_max_diff_win_series_prop,
      +val_max_diff_win_series_prop
    ),
    midpoint = 0
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_val_o_series_win_prop_over_expected
save_plot_for_slide(p_val_o_series_win_prop_over_expected)
```

@@

p_filt_val_o_series_win_prop_over_expected <- filt_val_o_win_prop |> 
  ggplot() +
  common_val_heatmap_layers() +
  heatmap_tile(
    .fill = 'diff_win_series_prop'
  ) +
  # geom_abline(
  #   color = 'white',
  #   linetype = 2,
  #   aes(
  #     slope = 1,
  #     intercept = 0
  #   )
  # ) +
  heatmap_text_layers(
    .df = filt_val_o_win_prop, 
    .f = o_series_win_over_expected_text_layer,
    .threshold = 0.1,
    .labeller = val_heatmap_div_labeller
  ) +
  val_swap_path() +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('brown'),
    high = extract_ggsci_mid_color('cyan'),
    mid = 'white',
    limits = c(
      -filt_val_max_diff_win_series_prop,
      +filt_val_max_diff_win_series_prop
    ),
    midpoint = 0
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_filt_val_o_series_win_prop_over_expected
save_plot_for_slide(p_val_o_series_win_prop_over_expected)
```

@@

@@

```{r}
#| label: altnerative-to-p=0.5-method-1
## p. 365 on https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf
theoretical_cod_series_length <- function(p) {
  6 * prob_of_series_lasting_n_games(n = 6, p = p, max_round = 11) +
    7 * prob_of_series_lasting_n_games(n = 7, p = p, max_round = 11) + 
    8 * prob_of_series_lasting_n_games(n = 8, p = p, max_round = 11) + 
    9 * prob_of_series_lasting_n_games(n = 9, p = p, max_round = 11) +
    10 * prob_of_series_lasting_n_games(n = 10, p = p, max_round = 11) +
    11 * prob_of_series_lasting_n_games(n = 11, p = p, max_round = 11)
}

cod_theoretical_series_lengths <- cod_ps |> 
  mutate(
    theoretical_series_length = map_dbl(p, theoretical_cod_series_length)
  )

cod_rounds_per_series <- cod_streaks_res$series |> 
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )

cod_theoretical_constant_p <- cod_theoretical_series_lengths |> 
  mutate(
    diff = theoretical_series_length - cod_rounds_per_series$actual
  ) |> 
  slice_min(abs(diff), n = 1) |> 
  pull(p)
cod_theoretical_constant_p
```

```{r}
#| label: altnerative-to-p=0.5-method-2
pluck_cod_n <- function(.n_rounds) {
  cod_streaks_res$series |>
    filter(n_rounds == .n_rounds) |> 
    pull(n_actual)
}

n6 <- pluck_cod_n(6)
n7 <- pluck_cod_n(7)
n8 <- pluck_cod_n(8)
n9 <- pluck_cod_n(9)
n10 <- pluck_cod_n(10)
n11 <- pluck_cod_n(11)
pm <- pmin(n6, n7, n8, n9, n10, n11)
n6adj <- round(10 * n6 / pm)
n7adj <- round(10 * n7 / pm)
n8adj <- round(10 * n8 / pm)
n9adj <- round(10 * n9 / pm)
n10adj <- round(10 * n10 / pm)
n11adj <- round(10 * n11 / pm)

maximize_cod_series_p <- function(p) {
  prob_of_series_lasting_n_games(n = 6, p = p, max_round = 11)^n6adj *
    prob_of_series_lasting_n_games(n = 7, p = p, max_round = 11)^n7adj * 
    prob_of_series_lasting_n_games(n = 8, p = p, max_round = 11)^n8adj * 
    prob_of_series_lasting_n_games(n = 9, p = p, max_round = 11)^n9adj *
    prob_of_series_lasting_n_games(n = 10, p = p, max_round = 11)^n10adj * 
    prob_of_series_lasting_n_games(n = 11, p = p, max_round = 11)^n11adj
}

cod_theoretical_constant_p <- cod_ps |> 
  mutate(
    theoretical_series_length = map_dbl(p, maximize_cod_series_p)
  )

cod_theoretical_constant_p <- cod_theoretical_constant_p |> 
  slice_max(theoretical_series_length, n = 1) |> 
  pull(p)
cod_theoretical_constant_p

m <- matrix(c(cod_streaks_res$series$n_actual, cod_streaks_res$series$n_expected), ncol = 2)
dimnames(m) <- list(n_rounds = as.character(6:11), kind = c('actual', 'expected'))
ch <- chisq.test(m)
# ch$expected
# chance_numbers <- tibble(
#   league = rep(c('mlb', 'nba', 'nhl'), each = 4),
#   i = rep(4:7, 3),
#   a = c(
#     c(0.1894, 0.215, 0.215, 0.3895),
#     c(0.1324, 0.25, 0.3382, 0.2794),
#     c(0.2532, 0.2405, 0.3038, 0.2025)
#   ),
#   e = rep(c(0.125, 0.25, 0.3125, 0.3125), 3)
# )
# chance_numbers |> filter(league == 'mlb') |> select(a, e) |> chisq.test()
```

>>>>

```{r}
#| label: setup-eval
library(tidyverse)
library(extrafont)
library(ggtext)
library(ggsci)
library(ggnewscale)
library(magick)
library(broom)
library(knitr)

@@

-   2020-2022 Call of Duty League (CDL) matches (qualifiers and majors)
-   4v4
-   First to 6 round wins
-   Teams swap sides (offense and defense) after every round
:::

@@

ggsci_cyan <- extract_ggsci_mid_color('cyan')
ggsci_brown <- extract_ggsci_mid_color('brown')
cod_max_diff_win_series_prop <- max(abs(cod_o_win_prop$diff_win_series_prop))
p_cod_o_series_win_prop_over_expected <- cod_o_win_prop |> 
  ggplot() +
  common_cod_heatmap_layers() +
  heatmap_tile(
    .fill = 'diff_win_series_prop'
  ) +
  heatmap_text_layers(
    .df = cod_o_win_prop, 
    .f = o_series_win_over_expected_text_layer,
    .threshold = 0.08,
    .labeller = cod_heatmap_div_labeller
  ) +
  scale_fill_gradient2(
    low = ggsci_brown,
    high = ggsci_cyan,
    mid = 'white',
    midpoint = 0,
    limits = c(
      -cod_max_diff_win_series_prop, 
      +cod_max_diff_win_series_prop
    )
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_cod_o_series_win_prop_over_expected
save_plot_for_slide(p_cod_o_series_win_prop_over_expected)
```

@@

val_max_diff_win_series_prop <- max(abs(val_o_win_prop$diff_win_series_prop))
p_val_o_series_win_prop_over_expected <- val_o_win_prop |> 
  ggplot() +
  common_val_heatmap_layers() +
  heatmap_tile(
    .fill = 'diff_win_series_prop'
  ) +
  heatmap_text_layers(
    .df = val_o_win_prop, 
    .f = o_series_win_over_expected_text_layer,
    .threshold = 0.2,
    .labeller = val_heatmap_div_labeller
  ) +
  val_swap_path() +
  scale_fill_gradient2(
    low = ggsci_brown,
    high = ggsci_cyan,
    mid = 'white',
    limits = c(
      -val_max_diff_win_series_prop,
      +val_max_diff_win_series_prop
    ),
    midpoint = 0
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_val_o_series_win_prop_over_expected
save_plot_for_slide(p_val_o_series_win_prop_over_expected)
```

@@

p_filt_val_o_series_win_prop_over_expected <- filt_val_o_win_prop |> 
  ggplot() +
  common_val_heatmap_layers() +
  heatmap_tile(
    .fill = 'diff_win_series_prop'
  ) +
  # geom_abline(
  #   color = 'white',
  #   linetype = 2,
  #   aes(
  #     slope = 1,
  #     intercept = 0
  #   )
  # ) +
  heatmap_text_layers(
    .df = filt_val_o_win_prop, 
    .f = o_series_win_over_expected_text_layer,
    .threshold = 0.1,
    .labeller = val_heatmap_div_labeller
  ) +
  val_swap_path() +
  scale_fill_gradient2(
    low = ggsci_brown,
    high = ggsci_cyan,
    mid = 'white',
    limits = c(
      -filt_val_max_diff_win_series_prop,
      +filt_val_max_diff_win_series_prop
    ),
    midpoint = 0
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_filt_val_o_series_win_prop_over_expected
save_plot_for_slide(p_val_o_series_win_prop_over_expected)
```

@@

## Should round win percentage in [COD SnD]{.cod} be modeled as constant?

::: fragment
[Chance (2019)](https://content.iospress.com/articles/journal-of-sports-analytics/jsa200422) finds that probability of winning a game that the title series in the MLB and NHL is **not** constant (at 50%)
:::

```{r}
#| label: chance-2019
chance_2019_leagues <- c('MLB World Series', 'NBA Finals', 'NHL Stanley Cup')

n_sports <- tibble(
  league = chance_2019_leagues,
  ## i think nba has to be 2019 instead of 2018 in order for the product of `n * prop_actual` to be a whole number
  n = c(2018 - 1923, 2019 - 1951, 2018 - 1939)
)

chance_sports <- tibble(
  league = rep(chance_2019_leagues, each = 4),
  games = rep(4:7, 3),
  prop_actual = c(
    c(0.1895, 0.2105, 0.2105, 0.3895),
    c(0.1324, 0.25, 0.3382, 0.2794),
    c(0.2532, 0.2278, 0.3165, 0.2025)
  ),
  prop_expected = rep(c(0.125, 0.25, 0.3125, 0.3125), 3)
)

sports <- chance_sports |> 
  inner_join(
    n_sports,
    by = 'league'
  ) |> 
  mutate(
    n_actual = round(n * prop_actual)
  )

generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

chi_sports <- sports |> 
  nest(data = -c(league)) |> 
  mutate(
    chi = map(data, ~chisq.test(.x$n_actual, p = .x$prop_expected) |> broom::tidy())
  ) |> 
  select(league, chi) |> 
  unnest(chi) |> 
  transmute(
    league, 
    label = generate_chi_label(statistic, p.value)
  )

sports |> 
  select(league, games, prop_expected, prop_actual) |> 
  mutate(
    across(
      starts_with('prop_'),
      scales::percent
    ),
    across(games, as.character)
  ) |> 
  pivot_wider(names_from = league, values_from = prop_actual) |> 
  bind_rows(
    bind_cols(
      tibble(games = 'Avg. # of games'),
      sports |> 
        distinct(games, prop_expected) |>
        summarize(prop_expected = as.character(round(sum(games * prop_expected), 2))),
      sports |> 
        group_by(league) |>
        summarize(z = as.character(round(sum(games * prop_actual), 2))) |> 
        pivot_wider(names_from = league, values_from = z)
    )
  ) |> 
  bind_rows(
    chi_sports |> 
      pivot_wider(names_from = league, values_from = label) |> 
      mutate(
        games = '$\\xi^2$ (p-value)'
      )
  ) |> 
  relocate(
    `Series ends in i games` = games,
    `Expected` = prop_expected
  ) |> 
  knitr::kable()
```

::: fragment
|         Series length | Expected | MLB World Series | NBA Finals | NHL Stanley Cup |
|----------------------:|---------:|-----------------:|-----------:|----------------:|
|                     4 |   12.50% |           18.95% |     13.24% |          25.32% |
|                     5 |   25.00% |           21.05% |     25.00% |          22.78% |
|                     6 |   31.25% |           21.05% |     33.82% |          31.65% |
|                     7 |   31.25% |           38.95% |     27.94% |          20.25% |
|                       |          |                  |            |                 |
|  **Avg. \# of games** |     5.81 |              5.8 |       5.76 |            5.47 |
| $\xi^2$ **(p-value)** |          |       8.7 (0.03) | 0.4 (0.94) |  13.6 (\<=0.01) |
:::

```{r}
#| label: cod_series_outcomes_chi

cod_rounds_per_series <- cod_streaks_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )
avg_actual_cod_rounds_per_series <- round(cod_rounds_per_series$actual, 2)
avg_expected_cod_rounds_per_series <- round(cod_rounds_per_series$expected, 2)

cod_series_outcomes_chi <- chisq.test(
  cod_streaks_res$series$n_actual, 
  p = cod_streaks_res$series$prop_expected
) |> 
  broom::tidy() |> 
  mutate(label = generate_chi_label(statistic, p.value)) |> 
  pull(label)
cod_series_outcomes_chi
```

```{r}
#| label: cod_series_outcomes_chi-eval
#| eval: true
avg_actual_cod_rounds_per_series <- 9.15
avg_expected_cod_rounds_per_series <- 9.29
cod_series_outcomes_chi <- '15.4 (<=0.01)'
```

::: fragment
-   In COD, the average number of rounds (**`r avg_actual_cod_rounds_per_series`**) is less than the expected value (**`r avg_expected_cod_rounds_per_series`**)
:::

::: fragment
-   A $\xi^2$ test for CoD demonstrates evidence against the constant probability null hypothesis for winning a given round in an SnD series (**`r cod_series_outcomes_chi`**)
:::

## Should round win percentage in [COD SnD]{.cod} be modeled as constant?

```{r}
#| label: alternative-to-p=0.5-method-1
## p. 365 on https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf
theoretical_cod_series_length <- function(p) {
  6 * prob_of_series_lasting_n_games(n = 6, p = p, max_round = 11) +
    7 * prob_of_series_lasting_n_games(n = 7, p = p, max_round = 11) + 
    8 * prob_of_series_lasting_n_games(n = 8, p = p, max_round = 11) + 
    9 * prob_of_series_lasting_n_games(n = 9, p = p, max_round = 11) +
    10 * prob_of_series_lasting_n_games(n = 10, p = p, max_round = 11) +
    11 * prob_of_series_lasting_n_games(n = 11, p = p, max_round = 11)
}

cod_theoretical_series_lengths1 <- cod_ps |> 
  mutate(
    theoretical_series_length = map_dbl(p, theoretical_cod_series_length),
    diff = theoretical_series_length - cod_rounds_per_series$actual
  )

cod_theoretical_constant_p1 <- cod_theoretical_series_lengths1 |> 
  slice_min(abs(diff), n = 1) |> 
  pull(p)

pretty_cod_theoretical_constant_p1 <- cod_theoretical_constant_p1 |> 
  scales::percent(accuracy = 0.01)
```

```{r}
#| label: alternative-to-p=0.5-method-1-eval
#| eval: true
pretty_cod_theoretical_constant_p1 <- '57.25%'
```

```{r}
#| label: cod-alternative-to-p=0.5-method-2
pluck_cod_n <- function(.n_rounds) {
  cod_streaks_res$series |>
    filter(n_rounds == .n_rounds) |> 
    pull(n_actual)
}

cod_n6 <- pluck_cod_n(6)
cod_n7 <- pluck_cod_n(7)
cod_n8 <- pluck_cod_n(8)
cod_n9 <- pluck_cod_n(9)
cod_n10 <- pluck_cod_n(10)
cod_n11 <- pluck_cod_n(11)

## This adjustment is something that is not done in the paper. I do it here to reduce
##   the magnitude of the exponents.
cod_pm <- pmin(cod_n6, cod_n7, cod_n8, cod_n9, cod_n10, cod_n11)
cod_n6adj <- round(10 * cod_n6 / cod_pm)
cod_n7adj <- round(10 * cod_n7 / cod_pm)
cod_n8adj <- round(10 * cod_n8 / cod_)
cod_n9adj <- round(10 * cod_n9 / cod_pm)
cod_n10adj <- round(10 * cod_n10 / cod_pm)
cod_n11adj <- round(10 * cod_n11 / cod_pm)

maximize_cod_series_p <- function(p) {
  prob_of_series_lasting_n_games(n = 6, p = p, max_round = 11)^cod_n6adj *
    prob_of_series_lasting_n_games(n = 7, p = p, max_round = 11)^cod_n7adj * 
    prob_of_series_lasting_n_games(n = 8, p = p, max_round = 11)^cod_n8adj * 
    prob_of_series_lasting_n_games(n = 9, p = p, max_round = 11)^cod_n9adj *
    prob_of_series_lasting_n_games(n = 10, p = p, max_round = 11)^cod_n10adj * 
    prob_of_series_lasting_n_games(n = 11, p = p, max_round = 11)^cod_n11adj
}

cod_theoretical_series_lengths2 <- cod_ps |> 
  mutate(
    theoretical_series_length = map_dbl(p, maximize_cod_series_p)
  )

cod_theoretical_constant_p2 <- cod_theoretical_series_lengths2 |> 
  slice_max(theoretical_series_length, n = 1) |> 
  pull(p)

pretty_cod_theoretical_constant_p2 <- cod_theoretical_constant_p2 |> 
  scales::percent(accuracy = 0.01)

best_cod_theoretical_series_lengths2 <- cod_theoretical_series_lengths2 |> 
  filter(p == cod_theoretical_constant_p2)

p_cod_theoretical_series_lengths2 <- cod_theoretical_series_lengths2 |> 
  ggplot() +
  aes(x = p, y = theoretical_series_length) +
  geom_line(
    size = 2,
    color = 'white'
  ) +
  geom_point(
    data = best_cod_theoretical_series_lengths2,
    color = ggsci_cyan,
    size = 6
  ) +
  geom_text(
    data = best_cod_theoretical_series_lengths2,
    color = ggsci_cyan,
    size = 13 / .pt,
    hjust = 0,
    aes(
      x = p + 0.02,
      label = sprintf('Maximum likilihood estimate (MLE) of\nconstant probability of better team winning (%s)', !!pretty_cod_theoretical_constant_p2)
    )
  ) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x = 'Constant probability of better team winning',
    y = 'Kernel of the multinomial likelihood'
  ) +
  theme(
    axis.text.y = element_blank(),
    panel.grid.major = element_blank()
  )
p_cod_theoretical_series_lengths2
save_plot_for_slide(p_cod_theoretical_series_lengths2)
```

```{r}
#| label: cod-alternative-to-p=0.5-method-2-eval
#| eval: true
pretty_cod_theoretical_constant_p2 <- '53.25%'
```

::: fragment
[Mosteller (1952)](https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf) suggests several ways to find the optimal constant probability of the "stronger" team winning each game in a series.
:::

::: fragment
-   Using the method of moments approach (Mosteller's method 1), `r pretty_cod_theoretical_constant_p1`
:::

::: fragment
-   Using the maximum likelihood approach (Mosteller's method 2), `r pretty_cod_theoretical_constant_p2`

![](plots/cod_theoretical_series_lengths2.png){fig-align="center" width="65%"}
:::

@@
